---
title: "3 Ways to Query Data"
output: html_notebook
---

You can query databases in with three methods:

* Pass through queries to DBI
* dplyr translation
* SQL code engines in R Markdown

# Setup

Load packages and create a generic plot function.

```{r}
library(tidyverse)
library(DBI)

plotFunction <- function(dat){
  ggplot(dat, aes(total, subscribe, color = year)) +
  geom_point() +
  xlab("Total contacts") +
  ylab("Term Deposit Subscriptions") +
  ggtitle("Contact volume")
}
```

# Connect

There are a few ways to connect to databases:

1. Connect to your database using a Data Source Name (DSN). 

Navigate to "Connections Pane" -> "New Connection" -> "Banks"

The Connections Pane creates a code snippet that generates the connection. Connect to the "Banks" DSN.


```{bash}
cat /etc/odbc.ini
```


```{r connection_dsn}
library(odbc)
con <- dbConnect(odbc::odbc(), "Banks")
```

Use the new RStudio object explorer to learn more about the `con` object. What version of the database are we using?

*ANSWER* 

```{r }
con@info[["db.version"]]

# OR

dbGetInfo(con)
```

2. Create a connection via arguments to the `dbConnect` function.

Try to Connect directly by filling out this template:

```{r connection, eval=FALSE}
con <- dbConnect(odbc::odbc(), 
        driver =    ,
        server =    ,
        uid =       ,
        pwd =       ,
        port =      ,
        database = 
)
```



You can also use the New Connection dialog under the RStudio Connections pane to create this type of connection. Navigate:

"Connections Pane" -> "New Connection" -> "SQLServer"

Bonus Question: Where does `Driver: SQLServer` get resolved?

*ANSWER*

```{bash}
cat /etc/odbcinst.ini
```

**Passwords**

- Store them in DSN or config files
- Store them in encrypted keychains
- Ask for them 

```{bash eval = FALSE}
pwd <- rstudioapi::askForPassword('Enter Password: ')
```


# DBI

Query your database using pass through SQL code to DBI.

```{r}
bank <- dbReadTable(con, "bank")

bank

bank_agg <- dbGetQuery(con,
"
select month_idx, year, month,
sum(case when term_deposit = 'yes' then 1.0 else 0.0 end) as subscribe,
count(*) as total
from bank
group by month_idx, year, month
")

bank_agg

plotFunction(bank_agg)
```

# dplyr

Query your database with dplyr code that gets translated to SQL.

```{r}
bank_tbl <- tbl(con, "bank")

# Challenge: Can you re-create the same query above with dplyr? 
```

*ANSWER* 

```{r}

bank_agg_tbl <- bank_tbl %>%
  group_by(month_idx, year, month) %>%
  summarise(
    subscribe = sum(ifelse(term_deposit == "yes", 1, 0)),
    total = n()
  )

bank_agg_tbl  %>%   collect()
```

It is actually easier to go the other way:

```{r}
show_query(bank_agg_tbl)
```

What does `collect` do? (Hint: What kind of object is `bank_agg_tbl`?)

```{r}
collect(bank_agg_tbl) %>%
  plotFunction(.)
```

# SQL

Query your database with pass through SQL in R Markdown. If you want to output the query to file use `output.var = "bank_agg_tbl2"`.

```{sql, connection=con}
select month_idx, year, month,
sum(case when term_deposit = 'yes' then 1.0 else 0.0 end) as subscribe,
count(*) as total
from bank
group by month_idx, year, month
```

```{sql, connection=con, output.var = "bank_agg_tbl2"}
select month_idx, year, month,
sum(case when term_deposit = 'yes' then 1.0 else 0.0 end) as subscribe,
count(*) as total
from bank
group by month_idx, year, month
```

```{r}
plotFunction(bank_agg_tbl2)
```


**Bonus** Can I go the other way? (Can I use a variable in R in my SQL code chunk?)

Answer: http://rmarkdown.rstudio.com/authoring_knitr_engines.html#sql

# Conclusion

You can use R Markdown, dplyr, DBI, and the RStudio connection pane to interact with your databases. For more information see [db.rstudio.com](http://db.rstudio.com).

